Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-03 TACO-paper-v2/TACO-Rebuttal.txt

We thank our reviewers (R1-R4). We appreciate the positive feedback and the
critical comments on our paper. Below we answers the various points arose by the
reviewers.

### 2014-09-03 TACO-paper-v2/coffee.tex

During the assembly phase, the solution of the PDE is approximated by
executing a problem-specific kernel over all cells, or elements, in the
discretized domain. In this work, we focus on relatively low order
finite element methods, in which an assembly kernel's working set is
usually small enough to fit the L1 cache. Low order methods are by no
means exotic: they are employed in a wide variety of fields, including
climate and ocean modeling, computational fluid dynamics, and structural
mechanics. High order methods such as the spectral element method
\cite{SPENCER} commonly require different algorithms to solve PDEs and
therefore are excluded from our study.

@@

The structure of a local assembly kernel is shown in
Figure~\ref{code:general-structure}. The inputs are a zero-initialized
two dimensional array used to store the element matrix, the element's
coordinates in the discretized domain, and coefficient fields, for
instance indicating the values of velocity or pressure in the
element. The output is the evaluated element matrix. The kernel body
can be logically split into three parts:
\begin{enumerate}
  \item Calculation of the Jacobian matrix, its determinant and its
    inverse required for the aforementioned change of coordinates from
    the reference element to the one being computed.
  % FIXME: this rather hand-wavy and not technically correct
  \item Definition of basis functions used, intuitively, to interpolate
    the contribution to the PDE solution over the element. The choice of
    basis functions is expressed in UFL directly by users. In the
    generated code, they are represented as global read-only two
    dimensional arrays (i.e., using \texttt{static const} in C) of
    double precision floats.
  \item Evaluation of the element matrix in an affine loop nest, in which
    the integration is performed.
\end{enumerate}
Table~\ref{table:map-name-letters} shows the variable names
we will use in the upcoming code snippets to refer to the various
kernel objects.

@@

We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, only a minimal performance gain is achieved. This changes instantly increasing $p$ or $f$. In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible. In such cases, the transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}. Interesting is the fact that generalized loop-invariant code motion was particularly invasive, with 23 temporaries generated and several redundancies discovered (see Section~\ref{sec:code-transf-licm}).

In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case $[f=2, p=2]$. However, we reiterate the fact that full-application speed ups rise proportionally with the amount of time spent in assembly and, therefore, with the complexity of the equation. By increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant. The choice of studying the static linear elasticity equation was to show that even relatively simple problems can be characterized by a large proportion of execution time spent in assembly.

@@

\paragraph{Generalizability to other domains} There are neither conceptual nor technical reasons for which our compiler could not be used outside of Firedrake. For example, integration with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly, would be relatively easy to achieve. More challenging is to assess the generality of the optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance. To answer these questions, we first need to go back to the origins of our compiler. The starting point of our work was the mathematical formulation of a local assembly operation, expressible as follows

@@

\item \textbf{Optimizing mathematical expressions.} Expression manipulation (e.g. simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality. The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture. Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domains involving numerical evaluation of complex mathematical expressions (e.g. electronic structure calculations in physics and quantum chemistry relying on tensor contractions~\cite{TCE}). In this context, we highlight three notable points.

@@

\item Associative arithmetic operators are the prerequisite for expression splitting. In essence, this transformation concerns resource-aware execution. It is indeed true that, in the context of COFFEE, expression splitting has been successfully applied to improve register pressure; nevertheless, the basic idea is far more general, which is re-scheduling (re-associating) operations to optimize for some generic parameters. It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by less stringent memory requirements (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}). In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also $\alpha$ could be completely extracted and evaluated in a separate kernel: this would avoid allocating space for $[a', b', c', ...]$ in a second kernel computing the rest of the expression. This may be particularly indicated for many-core architectures in which the available per-core memory is much smaller than that in traditional CPUs.

@@

\item \textbf{Code generation and applicability of the transformations.} All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time. This means that ``good'' code can be generated. For example, loop bounds can be made explicit, arrays can be statically initialized, and pointer aliasing is easily avoidable. Further, all of these factors contribute to the applicability and the effectiveness of some of our code transformations. For instance, knowing loop bounds allows both generating correct code when applying vector-register tiling (see Section~\ref{sec:op-vect}) and discovering redundant computation opportunities (see Section~\ref{sec:code-transf-licm}). A special case are padding and data alignment since they could be performed even at run-time if some values were not known at code generation time; theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.  
\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has become increasingly effective in a variety of codes. Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers, to the best of our knowledge. The outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization. In addition, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi experiments). Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that this could be more easily achieved in specific domains. The intuition is to map the memory access pattern on vector registers, and then exploiting in-register shuffling to minimize the traffic between memory and processor. By demonstrating the effectiveness of multi-loop vectorization in a real scenario, our research represents an incentive for studying this technique in a broader and systematic way. 

>>>>

During the assembly phase, the solution of the PDE is approximated by
executing a problem-specific kernel over all cells, or elements, in the
discretized domain. In this work, we focus on relatively low order finite
element methods, in which an assembly kernel's working set is usually small
enough to fit the L1 cache. Low order methods are by no means exotic: they
are employed in a wide variety of fields, including climate and ocean
modeling, computational fluid dynamics, and structural mechanics. The
efficient assembly of high order methods such as the spectral element
method \cite{SPENCER} requires a significantly different loop nest structure.
High order methods  therefore are excluded from our study.

@@

The structure of a local assembly kernel is shown in
Figure~\ref{code:general-structure}. The inputs are a zero-initialized
two dimensional array used to store the element matrix, the element's
coordinates in the discretized domain, and coefficient fields, for
instance indicating the values of velocity or pressure in the
element. The output is the evaluated element matrix. The kernel body
can be logically split into three parts:
\begin{enumerate}
  \item Calculation of the Jacobian matrix, its determinant and its
    inverse required for the aforementioned change of coordinates from
    the reference element to the one being computed.
  % FIXME: this rather hand-wavy and not technically correct
  \item Definition of basis functions used to interpolate fields at the
    quadrature points in the element. The choice of basis functions is
    expressed in UFL directly by users. In the generated code, they are
    represented as global read-only two dimensional arrays (i.e., using
    \texttt{static const} in C) of double precision floats.
  \item Evaluation of the element matrix in an affine loop nest, in which
    the integration is performed.
\end{enumerate}
Table~\ref{table:map-name-letters} shows the variable names
we will use in the upcoming code snippets to refer to the various
kernel objects.

@@

We first notice that, in the scenario ($f=1, p=1$), the assembly is
dominated by matrix insertion: despite the application of several
transformations, only a minimal performance gain is achieved. However if
either $p$ or $f$ is increased then the cost of \texttt{assembly} becomes
larger with respect to \texttt{solve}, and the matrix insertion cost becomes
negligible. In such cases, the transformations automatically applied by
COFFEE dramatically decrease the cost of \texttt{assembly}.  This results in
a significant overall speedup and a simulation which is now dominated by
\texttt{solve} time. It is interesting to note that
generalized loop-invariant code motion was particularly invasive in this
case, with 23 temporaries generated and several redundancies discovered (see
Section~\ref{sec:code-transf-licm}).

In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case ($f=2, p=2$). However, we reiterate the fact that full-application speed ups rise proportionally with the amount of time spent in assembly and, therefore, with the complexity of the equation. By increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant. The choice of studying the static linear elasticity equation was to show that even relatively simple problems can be characterized by a large proportion of execution time spent in assembly.

@@

\paragraph{Generalizability to other domains} There are neither conceptual
nor technical reasons which prevent our compiler being used in applications
other than Firedrake. For example, integration with the popular FEniCS
framework, the pioneer of automated code generation for finite element local
assembly, would be relatively easy to achieve. It is more challenging to
assess the generality of the optimization strategy: the extent to which COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance. To answer these questions, we first need to go back to the origins of our compiler. The starting point of our work was the mathematical formulation of a local assembly operation, expressible as follows

@@

\item \textbf{Optimizing mathematical expressions.} Expression manipulation
  (e.g. simplification, decomposition into sub-expressions) opens multiple
  semantically equivalent code generation opportunities, characterized by
  different trade-offs in parallelism, redundant computation, and data
  locality. The basic idea is to exploit properties of arithmetic operators,
  such as associativity and commutativity, to re-schedule the computation suitably for the underlying architecture. Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domains involving numerical evaluation of complex mathematical expressions (e.g. electronic structure calculations in physics and quantum chemistry relying on tensor contractions~\cite{TCE}). In this context, we highlight three notable points.

@@

\item Associative arithmetic operators are the prerequisite for expression
  splitting. In essence, this transformation concerns resource-aware
  execution. In the context of COFFEE, expression splitting has been
  successfully applied to improve register pressure. However, underlying
  idea of re-scheduling (re-associating) operations to optimize for some
  generic parameters is far more general. It could be used, for example, as
  a starting point to perform kernel fission; that is, splitting a kernel
  into multiple parts, each part characterized by less stringent memory
  requirements (a variant of this idea for non-affine loops in unstructured
  mesh applications has been adopted in~\cite{carlolcpc}). In
  Equation~\ref{eq:assembly-model}, for instance, not only can  any of the
  functions $\alpha$, $\beta$ and $\gamma$ be split (assuming they include
  associative operators), but $\alpha$ could be completely extracted and
  evaluated in a separate kernel. This would reduce the working set size of
  each of the kernel functions, an option which is particularly attractive for many-core architectures in which the available per-core memory is much smaller than that in traditional CPUs.

@@

\item \textbf{Code generation and applicability of the transformations.} All
  array sizes and loop bounds, like $n1$ and $n2$ in
  Equation~\ref{eq:assembly-model}, are known at code generation time. This
  means that ``good'' code can be generated. For example, loop bounds can be
  made explicit, arrays can be statically initialized, and pointer aliasing
  is easily avoidable. Further, all of these factors contribute to the
  applicability and the effectiveness of some of our code
  transformations. For instance, knowing loop bounds allows both generating
  correct code when applying vector-register tiling (see
  Section~\ref{sec:op-vect}) and discovering redundant computation
  opportunities (see Section~\ref{sec:code-transf-licm}). Padding and data
  alignment are spectial cases, since they could be performed at run-time if some values were not known at code generation time. Theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.  
\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has
  become increasingly effective in a variety of codes. However, to the best
  of our knowledge, multi-loop vectorization involving the loading and
  storing of data along a subset of the loops characterizing the iteration
  space (rather than just along the innermost loop), is not supported by
  available general-purpose and polyhedral compilers. The outer-product
  vectorization technique presented in this paper shows that two-loop
  vectorization can outperform standard auto-vectorization. In addition, we
  expect the performance gain to scale with the number of vectorized loops
  and the vector length (as demonstrated in  the Xeon Phi
  experiments). Although the automation of multi-loop vectorization in a
  general-purpose compiler is far from straightforward, especially if stencils are present, we believe that this could be more easily achieved in specific domains. The intuition is to map the memory access pattern on vector registers, and then exploit in-register shuffling to minimize the traffic between memory and processor. By demonstrating the effectiveness of multi-loop vectorization in a real scenario, our research represents an incentive for studying this technique in a broader and systematic way. 
